Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-23 ICLR25/formula.tex
formula
for
form
update

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
Improve\_max$^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have Improve\_max$^z(a)\geq 0$ as Sol\_max$_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so Sol\_max$_{\{a\}}^z(z) \leq$ Sol\_max$_\emptyset^z(z)$.

Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z}$ and 
Improve\_min$_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of Improve\_max$_a^z$.

@@

For all neurons $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.

%For one layer before the target layer, the formula is simple and most accurate. 
To estimate Improve\_max$_a^z$, we define Utility\_max$^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		Utility\_max^z(a) = |W_{az} \times (sol(\hat{a})- \ReLU(sol(a)))|.
	\end{align}

@@

\begin{proposition}
$0 \leq Improve\_max^z(a) \leq Utility\_\max^z(a)$. 
\end{proposition}

@@

Now, it is easy to check that $(sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as openning $a$ changes the value of $\hat{a}$ from
$sol(\hat{a})$ to $\ReLU(sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $sol'(z) \leq$ Sol\_max$_{\{a\}}^z(z)$. That is, 
$sol(z)-sol'(z) \geq sol(z) -$ Sol\_max$_{\{a\}}^z(z)$, i.e. 
$Improve\_max^z(a) \leq Utility\_\max^z(a)$.
\end{proof}

Thus, Utility\_max$^z(a)$ can be used to approximate 
Improve\_max$^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $sol(\hat{a})$ to its minimal value\dots which happens to be $sol(\hat{a})=ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have Utility\_max$^z(a)=0$, and thus Improve\_max$^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  Sol\_max$_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

@@

%To estimate Improve\_max$_a^z$ , 
Indeed, if $W_{bz}\leq0$, then the LP solver sets $sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $ReLU(b)$ according to Proposition \ref{LP}.
If $sol(b) < 0$, then we have a clip at $sol(\hat{b})=ReLU(b)=0$ and opening $a$ does not help. If $sol(b) > 0$, then we have $sol(\hat{b})=ReLU(b)=b$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $b$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $sol(\hat{b}) = sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

@@

We define a distance $D$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&D(\hat{a}) = sol(\hat{a}) - \ReLU(sol(a)) \geq 0\\
			&D(\hat{b}) =
		\begin{cases}
			sol(\hat{b}) - \min(\UB(b), sol(\hat{b}) - D(\hat{a}) k(\hat{b})), & \text{if }  W_{bz} > 0 \\
			sol(\hat{b}) - \max(\LB(b), 0, sol(\hat{b}) -D(\hat{a}) k(\hat{b})), & \text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

@@

$$ Utility\_max^z(a) = \sum_b W_{bz} D(\hat{b})$$

We proceed inductively in the same way to define Utility$\_max^z(a)$ for deeper neuron $a$.

@@

	First, $D(a) = \ReLU(sol(a))-sol(\hat{a})$.

To compute $D(b)$ for neurons $b$ in two layer before $z$, we define \begin{align}
	&u_0 = \max(\LB(b),\min(\UB(b),  sol(b)+D(a)W_{ab}))\\
	&u_1 = \begin{cases}
		\ReLU(u_0)+h(b)(\frac{\UB(c)u_0-\UB(b)\LB(b)}{\UB(b)-\LB(b)}-\ReLU(u_0)), & \text{if }\LB(b) < 0\\
	u_0, & \text{if }  \LB(b) \geq 0
	\end{cases}\\
	&D(b) = u_1-sol(\hat{b})
\end{align}

	To compute $D(c)$ for neurons $c$ in one layer before $z$, we define 
	\begin{align}
		&w_0 = \sum_b D(b)W_{bc}\\
		&w_1 = \min(\UB(c),sol(c)+w_0)\\		
		&D(c) =
		\begin{cases}
			w_1-sol({c}), & \text{if }W_{cz} > 0 \text{ and } \LB(c)\geq 0\\
		k(c)(w_1-sol({c})), & \text{if }W_{cz} > 0 \text{ and } \LB(c)< 0\\
		\ReLU(w_1)-sol(\hat{c})	, & \text{if }  W_{cz} < 0
		\end{cases}\\
		&D(z) = \sum_c D(c)W_{cz}\\
		&Utility\_max^z(a) = -D(z)
	\end{align}
\end{definition}

>>>>

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
$\Improve\_max^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_max^z(a)\geq 0$ as Sol\_max$_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so Sol\_max$_{\{a\}}^z(z) \leq$ Sol\_max$_\emptyset^z(z)$.
Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_min_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of $\Improve\_max_a^z$.

@@

For all neurons $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.
%For one layer before the target layer, the formula is simple and most accurate. 
To estimate $\Improve\_max_a^z$, we define $\Utility$\_max$^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		\Utility\_max^z(a) = |W_{az} \times (sol(\hat{a})- \ReLU(sol(a)))|.
	\end{align}

@@

\begin{proposition}
$0 \leq \Improve\_max^z(a) \leq \Utility\_\max^z(a)$. 
\end{proposition}

@@

Now, it is easy to check that $(sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as openning $a$ changes the value of $\hat{a}$ from
$sol(\hat{a})$ to $\ReLU(sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $sol'(z) \leq$ Sol\_max$_{\{a\}}^z(z)$. That is, 
$sol(z)-sol'(z) \geq sol(z) -$ Sol\_max$_{\{a\}}^z(z)$, i.e. 
$\Improve\_max^z(a) \leq Utility\_\max^z(a)$.
\end{proof}

Thus, Utility\_max$^z(a)$ can be used to approximate 
$\Improve\_max^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $sol(\hat{a})$ to its minimal value\dots which happens to be $sol(\hat{a})=\ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have Utility\_max$^z(a)=0$, and thus $\Improve\_max^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  Sol\_max$_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

@@

%To estimate $\Improve\_max_a^z$ , 
Indeed, if $W_{bz}\leq0$, then the LP solver sets $sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $\ReLU(b)$ according to Proposition \ref{LP}.
If $sol(b) < 0$, then we have a clip at $sol(\hat{b})=\ReLU(b)=0$ and opening $a$ does not help. If $sol(b) > 0$, then we have $sol(\hat{b})=\ReLU(b)=sol(b)$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $\hat{b}$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $sol(\hat{b}) = sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

@@

We define a difference $\Delta$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&\Delta(\hat{a}) = sol(\hat{a}) - \ReLU(sol(a)) \geq 0\\
			&\Delta(\hat{b}) =
		\begin{cases}
			sol(\hat{b}) - \min(\UB(b), sol(\hat{b}) - \Delta(\hat{a}) k(\hat{b})), & \text{if }  W_{bz} > 0 \\
			sol(\hat{b}) - \max(\LB(b), 0, sol(\hat{b}) -\Delta(\hat{a}) k(\hat{b})), & \text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

@@

$$ \Utility\_max^z(a) = \sum_b W_{bz} \Delta(\hat{b})$$

We proceed inductively in the same way to define $\Utility\_max^z(a)$ for deeper neuron $a$.

@@

	First, $\Delta(a) = \ReLU(sol(a))-sol(\hat{a})$.

To compute $\Delta(b)$ for neurons $b$ in two layer before $z$, we define \begin{align}
	&u_0 = \max(\LB(b),\min(\UB(b),  sol(b)+\Delta(a)W_{ab}))\\
	&u_1 = \begin{cases}
		\ReLU(u_0)+h(b)(\frac{\UB(c)u_0-\UB(b)\LB(b)}{\UB(b)-\LB(b)}-\ReLU(u_0)), & \text{if }\LB(b) < 0\\
	u_0, & \text{if }  \LB(b) \geq 0
	\end{cases}\\
	&\Delta(b) = u_1-sol(\hat{b})
\end{align}

	To compute $\Delta(c)$ for neurons $c$ in one layer before $z$, we define 
	\begin{align}
		&w_0 = \sum_b \Delta(b)W_{bc}\\
		&w_1 = \min(\UB(c),sol(c)+w_0)\\		
		&\Delta(c) =
		\begin{cases}
			w_1-sol({c}), & \text{if }W_{cz} > 0 \text{ and } \LB(c)\geq 0\\
		k(c)(w_1-sol({c})), & \text{if }W_{cz} > 0 \text{ and } \LB(c)< 0\\
		\ReLU(w_1)-sol(\hat{c})	, & \text{if }  W_{cz} < 0
		\end{cases}\\
		&\Delta(z) = \sum_c \Delta(c)W_{cz}\\
		&\Utility\_max^z(a) = -\Delta(z)
	\end{align}
\end{definition}

### 2024-09-23 ICLR25/valabstraction.tex
update

\iffalse
An extremely efficient algorithm that alleviates some of the inaccuracies is DeepPoly \cite{deeppoly}, also independently discovered as the CROWN algorithm \cite{crown}. Instead of fixed bounds for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$. For bounds $[\alpha,\beta]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. 
\fi

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a complete (but inefficient) method. 
Consider an unstable neuron $n \in[\alpha,\beta]$. The value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in $\{0,1\}$, using constants $(\alpha,\beta)$ with 4 constraints \cite{MILP}:
$\quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a)$.

For all $x \in [\alpha,\beta] \setminus 0$, there exists a unique solution $(a,\hat{x})$ that meets these constraints, with $\hat{x}=\ReLU(x)$ \cite{MILP}. Here, $a$ is 0 if $x < 0$, 1 if $x>0$, and can be either if $x=0$~\cite{MILP}. This encoding approach can be applied to every (unstable) ReLU node, and optimizing its value can help getting more accurate bounds. However, for networks with hundreds of {\em unstable} nodes, the resulting MILP formulation will contain numerous integer variables and generally bounds obtained will not be accurate, even using powerful commercial solvers such as Gurobi.

>>>>

%\iffalse
%An extremely efficient algorithm that alleviates some of the inaccuracies is DeepPoly \cite{deeppoly}, also independently discovered as the CROWN algorithm \cite{crown}. Instead of fixed bounds for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$. For bounds $[\UB,\LB]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)=\LB \frac{x_i-\alpha}{\LB-\UB}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. 
%\fi

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a complete (but inefficient) method. 
Consider an unstable neuron $n \in[\UB,\LB]$. The value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in $\{0,1\}$, using constants $(\UB,\LB)$ with 4 constraints \cite{MILP}:
$\quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \LB \cdot a \quad \wedge \quad \hat{x} \leq x-\UB \cdot (1-a)$.

For all $x \in [\UB,\LB] \setminus 0$, there exists a unique solution $(a,\hat{x})$ that meets these constraints, with $\hat{x}=\ReLU(x)$ \cite{MILP}. Here, $a$ is 0 if $x < 0$, 1 if $x>0$, and can be either if $x=0$~\cite{MILP}. This encoding approach can be applied to every (unstable) ReLU node, and optimizing its value can help getting more accurate bounds. However, for networks with hundreds of {\em unstable} nodes, the resulting MILP formulation will contain numerous integer variables and generally bounds obtained will not be accurate, even using powerful commercial solvers such as Gurobi.
